---
title: "Testing a cluster count with matched nulls"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing a cluster count with matched nulls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_mclust <- requireNamespace("mclust", quietly = TRUE)
```

## The question

A clustering method applied to continuous data will report some number of
clusters whether or not any exist. So when an analysis announces that a
dataset contains, say, four types, the number may describe the population, or
only the shape of the data under that method. `matchednull` separates the
two.

## The idea: a null twin

`copula_null()` builds a synthetic twin of your data that keeps everything
innocent and removes the one thing in question:

* every variable's distribution is preserved **exactly** (the twin reuses the
  real values, reshuffled);
* the correlation matrix is preserved to within sampling error;
* all remaining dependence is Gaussian, so the twin contains **no cluster
  structure by construction**.

If your pipeline finds as many clusters in the twins as in the real data,
those clusters were the data's shape, not its people.

```{r}
library(matchednull)

set.seed(1)
x <- matrix(rnorm(400 * 3), 400, 3) %*%
  chol(matrix(c(1, .5, .3, .5, 1, .4, .3, .4, 1), 3, 3))
twin <- copula_null(x)

all(sort(twin[, 1]) == sort(x[, 1]))   # margins: identical
round(cor(x) - cor(twin), 2)           # correlations: close
```

## The test

`matched_null_test()` takes your data and **your own pipeline**, wrapped as a
function that returns one number (typically the selected number of clusters).
It runs the identical pipeline on the real data and on `R` twins, and asks
whether the real answer stands out.

```{r, eval = has_mclust}
suppressPackageStartupMessages(library(mclust))
pick_k <- function(d) Mclust(d, G = 1:4, modelNames = "VVV", verbose = FALSE)$G

# 1. Typeless data: the test should stay quiet.
set.seed(7)
matched_null_test(x, pick_k, R = 30)
```

```{r, eval = has_mclust}
# 2. Two genuine types, hidden in the dependence structure
#    (identical margins, opposite correlation orientation).
set.seed(42)
z <- sample(2, 400, replace = TRUE)
X <- matrix(rnorm(400 * 4), 400, 4)
L1 <- chol(matrix(c(1, .85, .85, 1), 2, 2))
L2 <- chol(matrix(c(1, -.85, -.85, 1), 2, 2))
X[z == 1, 1:2] <- X[z == 1, 1:2] %*% L1
X[z == 1, 3:4] <- X[z == 1, 3:4] %*% L1
X[z == 2, 1:2] <- X[z == 2, 1:2] %*% L2
X[z == 2, 3:4] <- X[z == 2, 3:4] %*% L2

set.seed(7)
matched_null_test(X, pick_k, R = 30)
```

The first call returns *null-like*: whatever clustering the pipeline reports
is reproduced by twins with no types. The second returns *exceeds the null*:
the grouping lives in structure the twins cannot carry.

## Plugging in your own pipeline

`cluster_fn` can wrap anything: a k-means heuristic, a published typology's
exact workflow, or any scalar measure of clustering strength. The null is
defined at the level of the data, not the pipeline.

Two practical notes:

* **Set a seed before calling** — reproducibility is deliberately left to the
  caller.
* **Check the margins first.** The one regime the null cannot flag is types so
  separated that they are visible in the margins and covariance themselves
  (strongly bimodal variables); inspect the margins for pronounced
  multimodality before relying on the count test, and tie-break granular
  (Likert-type) scales before formal unimodality tests.

## Reference

The method, its positive controls, and its false-positive calibration are
described in the accompanying paper: *Types Without Taxa: A
Covariance-Matched-Null Multiverse Test of Categorical versus Continuous
Personality Structure*.
